---
title: "Evaluation of quadrat-related functions"
author: "Gabriel Arellano"
date: today
format:
  pdf:
    code-overflow: wrap
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    number-sections: true
    number-depth: 2
    colorlinks: true
    toccolor: blue
editor: visual
---

## Goal

There is a set of functions in the CTFS / fgeo functionality that seems related to handling quadrats and coordinates in ForestGEO plots. These are 80+ functions that seem to work around:

-   transformations
    -   local coordinates + quadrat info = global coordinates

    -   global coordinates + quadrat info = local coordinates

    -   quadrat code –\> quadrat location
-   splitting plot into pieces
    -   quadrats of standard size

    -   smaller quadrats

    -   hectares
-   selecting
    -   certain quadrats

    -   neighboring quadrats

    -   random quadrats

This code seems concentrated in the CTFS `quadfunc.r` script. That script may contain some extra functions not that directly related to these tasks. This is the evaluation of the whole script, and the dependencies it creates.

## Load functions

First, we declare the path to fgeo2 folder, and source useful auxiliary functions:

```{r}

# This is my local path. You may have to adapt this.
g <- regexpr("fgeo2", getwd())
path <- substr(getwd(), 1, g + attr(g, "match.length"))

# Source useful functionality:
source(paste0(path, "evaluation/0-code-to-source.r"))

# Other functionality:
library(igraph, quietly = TRUE)

```

This chunk of code will source all the scripts, something similar to the old "startCTFS" function, but following a different structure of folders:

```{r}

# Source the R scripts
f1 <- list.files(paste0(path, "01_utilities/R/old/"),
                 full.names = TRUE, pattern = "\\.R$", ignore.case = TRUE)
f2 <- list.files(paste0(path, "02_biomass/R/old/"),
                 full.names = TRUE, pattern = "\\.R$", ignore.case = TRUE)
f3 <- list.files(paste0(path, "03_spatial/R/old/"),
                 full.names = TRUE, pattern = "\\.R$", ignore.case = TRUE)
f4 <- list.files(paste0(path, "04_abundance_diversity/R/old/"),
                 full.names = TRUE, pattern = "\\.R$", ignore.case = TRUE)
f5 <- list.files(paste0(path, "05_demography_growth/R/old/"),
                 full.names = TRUE, pattern = "\\.R$", ignore.case = TRUE)
f6 <- list.files(paste0(path, "06_topography_habitat/R/old/"),
                 full.names = TRUE, pattern = "\\.R$", ignore.case = TRUE)
f7 <- list.files(paste0(path, "07_mapping_plotting/R/old/"),
                 full.names = TRUE, pattern = "\\.R$", ignore.case = TRUE)

ff <- c(f1, f2, f3, f4, f5, f6, f7)
for(f in ff) source(f)
```

Note that it works on (possibly) updated scripts. The history of changes is briefly being recorded in the "function_inventory" table. If you want to evaluate the structure of dependencies in the original code by Condit, change "new" by "old" in the path names above.

## Map dependencies

This maps all the mutual dependencies among the whole body of functions:

```{r}

# All functions in the target body of code:
all <- unique(unlist(lapply(ff, function(f) functions_in_script(f))))

# All of the dependencies among all functions in the global environment:
edges <- get_all_dependencies()

# Subset the dependencies for the target functions only:
keep <- edges$from %in% all | edges$to %in% all
edges <- edges[keep,,drop = FALSE]

# Proper graph representation, maybe useful:
g <- graph_from_data_frame(edges, directed = TRUE)

pdf(paste0(path, "evaluation/results/dependencies-", Sys.Date(), ".pdf"),
    height = 33, width = 33)

plot(g, layout = layout.fruchterman.reingold, 
     vertex.label = V(g)$name,
     vertex.size = 5,
     vertex.color= V(g)$color,
     vertex.frame.color= "white",
     vertex.label.color = "black",
     edge.width=E(g)$weight, 
     edge.color="black")

dev.off()
```

\
