---
title: "Evaluation of (probably) obsolete modeling functions"
author: "Gabriel Arellano"
date: today
format:
  pdf:
    code-overflow: wrap
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    number-depth: 2
    colorlinks: true
    toccolor: blue
    include-in-header:
      text: |
        \usepackage{enumitem}
        \usepackage{array}
        \usepackage{pdflscape}
        \newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
        \setlist{itemsep=2pt, topsep=2pt}
editor: visual
---

# Goal

The goal is to make the fgeo2 body of code lighter and easier to maintain by removing obsolete modeling functions. fgeo2 will focus on metrics, descriptors, and other reusable analytical primitives. Statistical modeling requires flexibility and is better handled with maintained, dedicated tools rather than project-specific model-fitting and MCMC infrastructure. This evaluation focuses on:

-   primarily, removal of modeling functions;
-   minimum edits to the remaining ones; and
-   identifying dependents somewhere else, so all code remains working.

# Candidate functions

An earlier function-by-function evaluation was prepared during the transition from the CTFS scripts to the `ctfs` and `fgeo` packages. The table was edited by Sean McMahon and Gabriel Arellano and discussed with Stuart Davies and Mauro Lepore. It is preserved in [`evaluation/old-evaluation-table.xlsx`](https://github.com/gabrielareto/fgeo2/blob/c053ae2740c942244b222f05893a190faca2187a/evaluation/old-evaluation-table.xlsx).

That old evaluation was taken into account together with an independent review of the current source, tests, dependencies, scientific role, and maintained alternatives. The independent review deliberately looked for reasons to retain the code rather than treating the previous recommendation as authoritative.

The table below is the consolidated evaluation. It gives one operational decision for each function: **remove**, **review**, or **keep**. The notes combine the old evaluation, the independent review, and evidence from the [CTFS development notes](https://github.com/forestgeo/ctfs/blob/cc32c7c2a3959ef60b13144f956e99d3a0515322/NEWS.md), [growth tests](https://github.com/forestgeo/ctfs/blob/cc32c7c2a3959ef60b13144f956e99d3a0515322/tests/testthat/test-growth_vs_dbh.R), and [legacy regression tests](https://github.com/forestgeo/ctfs/blob/cc32c7c2a3959ef60b13144f956e99d3a0515322/tests/testthat/test-replace_subset.R). Candidates are ordered by module and script so the source is easy to locate; module `01` means `01_utilities/R/new` and module `05` means `05_demography_growth/R/new`.

```{r}
#| echo: false
#| results: asis

candidate_table <- read.delim(
  text = "Module|Script|Function|Notes|Decision
01|utilities-lmerBayes.r|lmerBayes|The old review recommended deprecating the complete block. Bespoke MCMC lacks explicit priors and modern diagnostics; its disabled regression test checked only component names. Archive the model and API description.|remove
01|utilities-lmerBayes.r|lmerBayes.hyperllike.sigma|Sampler-specific covariance update; preserve an intended covariance prior only if documented.|remove
01|utilities-lmerBayes.r|lmerBayes.hyperllike.mean|Apparently unused generic likelihood; preserve the formula only if needed.|remove
01|utilities-lmerBayes.r|full.likelihood.lmerBayes|Likelihood accounting is questionable; preserve a formal factorization only for reproducibility.|remove
01|utilities-lmerBayes.r|llike.fixef.lmer|Thin adapter used only by the obsolete sampler.|remove
01|utilities-lmerBayes.r|llike.model.lmer|Incomplete observation likelihood with interactive failure behavior; record error families if migrating.|remove
01|utilities-lmerBayes.r|residual.llike.lmerBayes|Residual adapter used only by the obsolete sampler.|remove
01|utilities-lmerBayes.r|badSD|Trivial positivity check with no independent value.|remove
01|utilities-lmerBayes.r|arrangeParam.llike.2D|Parameter-packing helper tied to legacy sampler internals.|remove
01|utilities-lmerBayes.r|arrangeParam.Gibbs.2D|Parameter-packing helper tied to legacy Gibbs blocks.|remove
01|utilities-lmerBayes.r|saveParamFile|Ad hoc chain persistence; document the legacy format only if archived draws must be read.|remove
01|utilities-lmerBayes.r|restoreParamFile|Legacy restoration logic; create a one-time converter only if old fits require it.|remove
01|utilities-lmerBayes.r|resetParam|Reset plumbing used only by the obsolete persistence design.|remove
01|utilities-lmerBayes.r|summaryMCMC|Lacks modern chain diagnostics; preserve output names only for a legacy converter.|remove
01|utilities-lmerBayes.r|recalculate.lmerBayesllike|Broken defaults and mismatched arguments make reconstruction unreliable.|remove
01|utilities-lmerBayes.r|covTocorr|Duplicates stats::cov2cor().|remove
01|utilities-modelBayes.r|modelBayes|The old review recommended deprecating the complete block. Bespoke adaptive MCMC lacks explicit priors and diagnostics; a maintenance attempt still failed in debugging. Archive callback conventions only if useful.|remove
01|utilities-modelBayes.r|residual.llike.modelBayes|Internal residual adapter with no independent scientific role.|remove
01|utilities-modelBayes.r|summaryModelMCMC|Incomplete legacy summaries without modern convergence checks.|remove
01|utilities-statistics.r|regression.Bayes|Narrow conjugate regression with implicit priors and fragile inputs.|remove
01|utilities-statistics.r|Gibbs.regslope|Legacy sampler block with implicit prior and direct matrix inversion.|remove
01|utilities-statistics.r|Gibbs.regsigma|Legacy inverse-gamma update coupled only to regression.Bayes().|remove
01|utilities-statistics.r|Gibbs.normalmean|Generic conjugate update with implicit assumptions and no standalone use.|remove
01|utilities-statistics.r|Gibbs.normalvar|Generic update with an arbitrary zero-variance fallback.|remove
01|utilities-statistics.r|model.xy|Adaptive MCMC has implicit priors and inconsistent optional-SD handling; archive equations if needed.|remove
01|utilities-statistics.r|arrangeParam.llike|Internal parameter-packing helper with a faulty NULL path.|remove
01|utilities-statistics.r|arrangeParam.Gibbs|Internal packing helper used by legacy Gibbs blocks.|remove
01|utilities-statistics.r|llike.GaussModel|Generic likelihood used only by candidate modeling code.|remove
01|utilities-statistics.r|llike.GaussModelSD|Legacy heteroscedastic likelihood; preserve the variance specification only if reused.|remove
01|utilities-statistics.r|BadParam|Default validator always returns false and adds no protection.|remove
01|utilities-statistics.r|graph.modeldiag|Broken model-specific diagnostic plotting.|remove
01|utilities-statistics.r|metrop1step|Adaptive Metropolis step lacks safe adaptation, diagnostics, and robust failure handling.|remove
01|utilities-statistics.r|metrop1step.discrete|Untested discrete Metropolis step with no independent users.|remove
01|utilities-statistics.r|testmcmcfunc|Explicitly described as no longer used.|remove
05|demogchange-abund.fit.CTFS.r|model.littleR.Gibbs|Previously recommended for deprecation as paper-specific. Before deciding, map equations and priors to the paper and tutorial, correct known indexing and likelihood defects, and create a small simulated fixture with expected summaries. A historical test took about 20 minutes and no active automated test remains; do not retain the sampler unchanged.|review
05|demogchange-abund.fit.CTFS.r|full.abundmodel.llike|Previously recommended for deprecation as paper-specific. Reconstruct the generative factorization, verify that each density is counted once, and compare it with the paper and tutorial before archiving or reimplementation.|review
05|demogchange-abund.fit.CTFS.r|prob.N1|Previously recommended for deprecation as paper-specific. Validate the demographic observation equation and edge cases, including zero and small abundance, against the paper and simulated values.|review
05|demogchange-abund.fit.CTFS.r|spmean.mort.abundGibbs|Previously recommended for deprecation as paper-specific. Verify the mortality-abundance conditional density and parameter indexing against the paper and simulated data; record expected values.|review
05|demogchange-abund.fit.CTFS.r|hyper.abundGibbs|Sampler-only hyperparameter density; record prior choices in the model specification.|remove
05|demogchange-abund.fit.CTFS.r|hyper.mortGibbs|Sampler-only mortality hyperparameter block; record prior choices in the specification.|remove
05|demogchange-abund.fit.CTFS.r|bad.asympower.param|Trivial legacy constraint; record its valid parameter domain with the model specification.|remove
05|demogchange-abund.fit.CTFS.r|bad.asymexp.param|Trivial legacy constraint; record its valid parameter domain with the model specification.|remove
05|demogchange-abund.fit.CTFS.r|fitSeveralAbundModel|Repeats a slow, unreliable sampler without adding scientific capability. A historical abundance-model test took about 20 minutes and no active automated test remains.|remove
05|demogchange-abund.fit.CTFS.r|graph.abundmodel|Model-specific plotting; archive figure intent and variable definitions if needed.|remove
05|demogchange-abund.fit.CTFS.r|find.xaxis.hist|Minor histogram-axis helper with no independent value.|remove
05|demogchange-abund.fit.CTFS.r|abundmodel.fit|Sampler-specific summary helper with a broken confidence-interval branch.|remove
05|demogchange-demogChange.r|lmerMortLinear|Standard complementary-log-log hazard model; preserve the equation and a migration example.|remove
05|demogchange-demogChange.r|lmerMortFixedTime|Standard category-specific hazard model; preserve factor and time-offset conventions.|remove
05|growth-growthfit.bin.r|extract.growthdata|Previously grouped with growth-bin deprecation, but data preparation fits fgeo2 and has historical regression coverage. Keep the capability; repair AGB coupling, transformations, naming, and returnfull, and port a compact fixture.|keep
05|growth-growthfit.bin.r|run.growthbin.manyspp|Dataset-specific batch wrapper around failed models; its historical tutorial call failed and remains commented out in the formal test. Preserve only the intended result schema.|remove
05|growth-growthfit.bin.r|run.growthfit.bin|Orchestrates obsolete fitting engines; its historical tutorial call failed and remains commented out in the formal test. It has no independent scientific logic.|remove
05|growth-growthfit.bin.r|growth.flexbin|The old review proposed deprecating the growth-bin block. This unmaintained breakpoint fit failed in historical tutorial testing; preserve the piecewise-growth and error specification, not the engine.|remove
05|growth-growthfit.bin.r|llike.linearbin.optim|Likelihood adapter used only by the obsolete flexible-bin optimizer.|remove
05|growth-growthfit.bin.r|defineBinBreaks|Model-specific starting heuristic; preserve bin-width and sample rules only if later justified.|remove
05|growth-growthfit.bin.r|defineSDpar|Arbitrary model-specific starting values.|remove
05|growth-growthfit.bin.r|enoughSamplePerBin|Validator used only by the deprecated fitting chain; document the rule only if justified.|remove
05|growth-growthfit.bin.r|wideEnoughBins|Model-specific validator; document the minimum-width rule only if justified.|remove
05|growth-growthfit.bin.r|bad.binparam|Constraint wrapper tied to the obsolete breakpoint parameterization.|remove
05|growth-growthfit.bin.r|bad.binsdpar|Constraint wrapper tied to the obsolete residual-SD parameterization.|remove
05|growth-growthfit.bin.r|calculateBinModel.BIC|RSS-based calculation is not aligned with the fitted heteroscedastic likelihood.|remove
05|growth-growthfit.bin.r|calculateBinModel.DIC|Questionable reconstructed likelihood; preserve historical values only as metadata.|remove
05|growth-growthfit.bin.r|calculateBinModel.AIC|Implemented expression is not the standard AIC formula.|remove
05|growth-growthfit.bin.r|calculateBinModel.bestpred|Parameter and draw dimensions are interchanged, making predictions unreliable.|remove
05|growth-growthfit.bin.r|assembleBinOutput|Expects inconsistent fields and depends on broken upstream output.|remove
05|growth-growthfit.graph.r|graph.growthmodel.spp|Wrapper tied to obsolete model objects; preserve desired panels only as specifications.|remove
05|growth-growthfit.graph.r|graph.growthmodel|Model-specific plotting with no unique scientific computation.|remove
05|growth-growthfit.graph.r|overlay.growthbinmodel|Overlay helper tied to deprecated parameter and prediction structures.|remove
05|growth-growthfit.graph.r|compare.growthbinmodel|Uses the invalid AIC calculation and references the wrong object.|remove
05|growth-growthfit.graph.r|graph.outliers.spp|Previously treated as a low-priority growth graphic. Prototype a model-free record-difference table keyed by explicit plot, census, and tree identifiers, and test whether it adds useful measurement QC; keep only the extracted operation if it does.|review
05|growth-growthfit.graph.r|graph.outliers|Batch plotting wrapper; rebuild only if the extracted QC workflow proves useful.|remove
05|growth-growthfit.graph.r|binGraphSampleSpecies|Sampling and plotting wrapper tied to obsolete model outputs.|remove
05|growth-growthfit.graph.r|binGraphManySpecies.Panel|Untested panel helper tied to obsolete model outputs.|remove
05|growth-growthfit.graph.r|binGraphManySpecies|Top-level graphics wrapper inherits broken fitting and comparison dependencies.|remove",
  sep = "|",
  quote = "",
  colClasses = "character",
  check.names = FALSE
)

candidate_functions <- candidate_table$Function
removal_functions <- candidate_table$Function[
  candidate_table[["Decision"]] == "remove"
]
candidate_table_display <- candidate_table
names(candidate_table_display)[1] <- "Mod."

candidate_table_latex <- knitr::kable(
  candidate_table_display,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
  row.names = FALSE
)

candidate_table_latex <- sub(
  "\\begin{longtable}{lllll}",
  paste0(
    "\\begin{longtable}{@{}",
    "L{0.8cm}L{3cm}L{4.8cm}L{10.1cm}L{1.6cm}",
    "@{}}"
  ),
  candidate_table_latex,
  fixed = TRUE
)

cat(
  "\\begin{landscape}\n",
  "\\footnotesize\n",
  "\\setlength{\\tabcolsep}{3pt}\n",
  candidate_table_latex,
  "\n\\end{landscape}\n",
  sep = ""
)
```

## Functions requiring review

Before changing the code, review these functions and watch their internal, tutorial, test, and external dependents explicitly:

-   **Abundance model:** `model.littleR.Gibbs()`, `full.abundmodel.llike()`, `prob.N1()`, and `spmean.mort.abundGibbs()`. Compare the equations, priors, and parameterization with the publication and tutorial; reconstruct the likelihood factorization; document the known indexing and likelihood defects; and create a small simulated fixture with expected intermediate and final values. Then decide whether to archive only the scientific specification or reimplement it in a maintained framework. The current sampler should not be retained unchanged.
-   **Measurement QC:** `graph.outliers.spp()`. Separate the comparison of full and trimmed growth records from model fitting and plotting. Prototype a record-level table with explicit plot, census, and tree identifiers, test it on a compact fixture, and retain only that model-free operation if it provides useful quality control.

The single **keep** decision is `extract.growthdata()`: preserve its data-preparation role, repair the issues listed in the table, and port a compact regression fixture. The other 67 candidates are marked **remove**, subject to the dependency checks below.

# Due diligence for dependents and dependencies

This due diligence has two main goals:

1.  identify any function outside the candidate set that would break if a candidate were removed; and
2.  identify non-candidate functions that may no longer be necessary if the functions marked **remove** are removed.

The dependency structure below uses every R script in the `R/new` directory of all seven fgeo2 modules. The exact pre-work snapshot is commit [`040b73e`](https://github.com/gabrielareto/fgeo2/commit/040b73e0cd9a0d55d64c1639b22159d5c6cbd20e).

```{r}

path <- normalizePath(system2("git", c("rev-parse", "--show-toplevel"),
                              stdout = TRUE)[1])

library(igraph, quietly = TRUE)
source(file.path(path, "evaluation/0-code-to-source.r"))

pre_work_commit <- "040b73e0cd9a0d55d64c1639b22159d5c6cbd20e"

modules <- c(
  "01_utilities", "02_biomass", "03_spatial",
  "04_abundance_diversity", "05_demography_growth",
  "06_topography_habitat", "07_mapping_plotting"
)

all_files <- system2(
  "git",
  c("-C", shQuote(path), "ls-tree", "-r", "--name-only", pre_work_commit),
  stdout = TRUE
)

module_pattern <- paste(modules, collapse = "|")
scripts <- all_files[
  grepl(paste0("^(", module_pattern, ")/R/new/.*\\.[Rr]$"), all_files)
]

code_env <- new.env(parent = .GlobalEnv)

for(file in scripts)
{
  code <- system2(
    "git",
    c("-C", shQuote(path), "show", paste0(pre_work_commit, ":", file)),
    stdout = TRUE
  )

  eval(parse(text = paste(code, collapse = "\n")), envir = code_env)
}

objects <- mget(
  ls(code_env, all.names = TRUE),
  envir = code_env,
  inherits = FALSE
)

all_functions <- names(objects)[vapply(objects, is.function, logical(1))]

edges <- get_all_dependencies(envir = code_env)
edges <- edges[, c("from_function", "to_function"), drop = FALSE]
colnames(edges) <- c("from", "to")

candidates <- intersect(candidate_functions, all_functions)
removal_candidates <- intersect(removal_functions, all_functions)
missing_candidates <- setdiff(candidate_functions, all_functions)

if(length(missing_candidates))
  stop("Candidate functions not found: ", paste(missing_candidates, collapse = ", "))

```

## Dependents that could break

For each candidate, the following code follows dependency arrows through all levels and retains only dependents outside the candidate set. These are the functions that would require attention before removal.

```{r}
#| results: asis

dependent_rows <- list()

for(candidate in candidates)
{
  direct_dependents <- unique(edges$to[edges$from == candidate])
  direct_dependents <- setdiff(direct_dependents, candidates)

  all_dependents <- downstream_dependents(candidate, edges)
  all_dependents <- setdiff(all_dependents, candidates)

  if(length(all_dependents) == 0)
    next

  dependent_rows[[length(dependent_rows) + 1L]] <- data.frame(
    Candidate = candidate,
    `Direct dependents outside candidates` = paste(direct_dependents, collapse = ", "),
    `Dependents outside candidates at any level` = paste(all_dependents, collapse = ", "),
    check.names = FALSE
  )
}

if(length(dependent_rows) == 0)
{
  cat("**Result:** No function outside the candidate set depends on a candidate in the pre-work snapshot.\n")
} else {
  dependent_table <- do.call(rbind, dependent_rows)
  dependent_table_latex <- knitr::kable(
    dependent_table,
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE,
    row.names = FALSE
  )

  dependent_table_latex <- sub(
    "\\begin{longtable}{lll}",
    "\\begin{longtable}{@{}L{4cm}L{9cm}L{9cm}@{}}",
    dependent_table_latex,
    fixed = TRUE
  )

  cat(
    "\\begin{landscape}\n",
    "\\footnotesize\n",
    "\\setlength{\\tabcolsep}{4pt}\n",
    dependent_table_latex,
    "\n\\end{landscape}\n",
    sep = ""
  )
}

```

## Dependencies that may become unnecessary

This check follows dependencies upstream through all levels. A non-candidate dependency is marked as potentially unnecessary when it belongs to the dependency chain of functions marked **remove** and has no downstream use outside that potential removal set. Functions marked **keep** or **review** prevent a shared dependency from receiving that conclusion.

```{r}
#| results: asis

candidate_dependencies <- character()

for(candidate in removal_candidates)
  candidate_dependencies <- c(
    candidate_dependencies,
    upstream_dependencies(candidate, edges)
  )

candidate_dependencies <- unique(candidate_dependencies)
candidate_dependencies <- setdiff(candidate_dependencies, candidates)
potential_removal_set <- c(removal_candidates, candidate_dependencies)

dependency_rows <- list()

for(dependency in candidate_dependencies)
{
  all_users <- downstream_dependents(dependency, edges)
  removal_users <- intersect(all_users, removal_candidates)
  other_users <- setdiff(all_users, potential_removal_set)

  conclusion <- "Still required elsewhere"
  if(length(other_users) == 0)
    conclusion <- "Potentially unnecessary"

  dependency_rows[[length(dependency_rows) + 1L]] <- data.frame(
    Dependency = dependency,
    `Functions marked remove downstream` = paste(removal_users, collapse = ", "),
    `Other downstream functions` = if(length(other_users)) paste(other_users, collapse = ", ") else "None detected",
    Conclusion = conclusion,
    check.names = FALSE
  )
}

if(length(dependency_rows) == 0)
{
  cat("**Result:** No non-candidate fgeo2 dependencies were detected.\n")
} else {
  dependency_table <- do.call(rbind, dependency_rows)
  dependency_table <- dependency_table[
    order(
      dependency_table$Conclusion != "Potentially unnecessary",
      dependency_table$Dependency
    ),
  ]

  dependency_table_latex <- knitr::kable(
    dependency_table,
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE,
    row.names = FALSE
  )

  dependency_table_latex <- sub(
    "\\begin{longtable}{llll}",
    "\\begin{longtable}{@{}L{3.4cm}L{8.1cm}L{8.1cm}L{2.5cm}@{}}",
    dependency_table_latex,
    fixed = TRUE
  )

  cat(
    "\\begin{landscape}\n",
    "\\footnotesize\n",
    "\\setlength{\\tabcolsep}{4pt}\n",
    dependency_table_latex,
    "\n\\end{landscape}\n",
    sep = ""
  )
}

```

# Current evaluation and decisions

The consolidated evaluation recommends removing 67 of the 73 candidates, keeping and repairing `extract.growthdata()`, and completing the specified review for the five functions highlighted above. The main scientific qualification is the abundance model: its equations may be distinctive and worth preserving, but the current sampler and likelihood implementation should not be maintained as production software without reconstruction and validation.

General Bayesian, hierarchical, segmented, and model-comparison tasks are better handled by maintained tools such as Stan, `brms`, NIMBLE, `segmented`, `mcp`, and `loo`. fgeo2 should preserve only reusable data preparation, clearly justified measurement-QC operations, and any scientific specifications needed for reproducibility or future reimplementation. The legacy implementations remain available under the `old` directories and in repository history.

# Package structure after removals

The graph is intentionally deferred until after the removals. Once the work is complete, replace the post-work commit and date below and enable the chunk. The graph will use every top-level function in the seven `R/new` directories, including functions with no detected internal dependency edges.

```{r}
#| eval: false
#| out-width: "100%"

post_work_commit <- "REPLACE_WITH_POST_WORK_COMMIT"
post_work_date <- as.Date("YYYY-MM-DD")

post_all_files <- system2(
  "git",
  c("-C", shQuote(path), "ls-tree", "-r", "--name-only", post_work_commit),
  stdout = TRUE
)

post_scripts <- post_all_files[
  grepl(paste0("^(", module_pattern, ")/R/new/.*\\.[Rr]$"), post_all_files)
]

post_code_env <- new.env(parent = .GlobalEnv)

for(file in post_scripts)
{
  code <- system2(
    "git",
    c("-C", shQuote(path), "show", paste0(post_work_commit, ":", file)),
    stdout = TRUE
  )

  eval(parse(text = paste(code, collapse = "\n")), envir = post_code_env)
}

post_objects <- mget(
  ls(post_code_env, all.names = TRUE),
  envir = post_code_env,
  inherits = FALSE
)

post_functions <- names(post_objects)[vapply(post_objects, is.function, logical(1))]

post_edges <- get_all_dependencies(envir = post_code_env)
post_edges <- post_edges[, c("from_function", "to_function"), drop = FALSE]
colnames(post_edges) <- c("from", "to")

post_vertices <- data.frame(name = post_functions)
g.after <- graph_from_data_frame(
  post_edges,
  directed = TRUE,
  vertices = post_vertices
)

remaining_candidates <- intersect(candidate_functions, post_functions)
post_downstream <- character()

for(candidate in remaining_candidates)
  post_downstream <- c(post_downstream, downstream_dependents(candidate, post_edges))

post_downstream <- unique(post_downstream)
post_downstream <- setdiff(post_downstream, remaining_candidates)

V(g.after)$color <- "gray80"
V(g.after)$color[V(g.after)$name %in% post_downstream] <- "yellow"
V(g.after)$color[V(g.after)$name %in% remaining_candidates] <- "orange"

post_work_pdf <- file.path(
  path,
  "evaluation/results",
  paste0(
    "fgeo2-dependencies-", substr(post_work_commit, 1, 7), "-",
    format(post_work_date, "%Y-%m-%d"), ".pdf"
  )
)

dir.create(dirname(post_work_pdf), recursive = TRUE, showWarnings = FALSE)

set.seed(1)

pdf(post_work_pdf, height = 33, width = 33)
par(mar = c(0, 0, 2, 0))

plot(
  g.after,
  layout = layout.fruchterman.reingold,
  vertex.label = V(g.after)$name,
  vertex.size = 6,
  vertex.color = V(g.after)$color,
  vertex.frame.color = "white",
  vertex.label.color = "black",
  vertex.label.cex = 2,
  edge.color = "gray45",
  edge.width = 0.7,
  edge.arrow.size = 0.3
)

legend(
  "topleft",
  legend = c(
    "Candidate remaining",
    "Downstream dependent",
    "Other function"
  ),
  pt.bg = c("orange", "yellow", "gray80"),
  pch = 21,
  pt.cex = 2,
  bty = "n"
)

dev.off()

post_work_pdf_relative <- file.path("results", basename(post_work_pdf))
knitr::include_graphics(post_work_pdf_relative)

```
