---
title: "Evaluation of (probably) obsolete modeling functions"
author: "Gabriel Arellano"
date: today
format:
  pdf:
    code-overflow: wrap
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    number-depth: 2
    colorlinks: true
    toccolor: blue
    include-in-header:
      text: |
        \usepackage{enumitem}
        \usepackage{array}
        \usepackage{pdflscape}
        \newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
        \setlist{itemsep=2pt, topsep=2pt}
editor: visual
---

# Goal

The goal is to make the fgeo2 body of code lighter and easier to maintain by removing obsolete modeling functions. fgeo2 will focus on metrics, descriptors, and other reusable analytical primitives. Statistical modeling requires flexibility and is better handled with maintained, dedicated tools rather than project-specific model-fitting and MCMC infrastructure. This evaluation focuses on:

-   primarily, removal of modeling functions;
-   minimum edits to the remaining ones; and
-   identifying dependents somewhere else, so all code remains working.

# Candidates for removal

As a starting point, I will use an earlier function-by-function evaluation that was prepared during the transition from the CTFS scripts to the `ctfs` and `fgeo` packages. The table was edited by Sean McMahon and Gabriel Arellano and discussed with Stuart Davies and Mauro Lepore. That table is preserved in [`evaluation/old-evaluation-table.xlsx`](https://github.com/gabrielareto/fgeo2/blob/c053ae2740c942244b222f05893a190faca2187a/evaluation/old-evaluation-table.xlsx).

The table below groups the review into three practical categories:

-   **Strong:** explicit recommendation to deprecate the complete block, or an explicitly paper-specific implementation marked for deprecation.
-   **Weak:** “try to deprecate” or low priority, without a firm recommendation.
-   **After checks:** removal appeared appropriate, but the old review suspected that other functions might depend on the code.

The candidates are ordered by module and script so the source is easy to locate. Module `01` means `01_utilities/R/new`; module `05` means `05_demography_growth/R/new`, etc.

```{r}
#| echo: false
#| results: asis

candidate_table <- read.delim(
  text = "Module|Script|Function|Old recommendation|Summary of old decision and notes
01|utilities-lmerBayes.r|lmerBayes|Strong|Explicit recommendation to deprecate the complete lmerBayes block.
01|utilities-lmerBayes.r|lmerBayes.hyperllike.sigma|Strong|Explicit recommendation to deprecate the complete lmerBayes block.
01|utilities-lmerBayes.r|lmerBayes.hyperllike.mean|Strong|Explicit recommendation to deprecate the complete lmerBayes block.
01|utilities-lmerBayes.r|full.likelihood.lmerBayes|Strong|Explicit recommendation to deprecate the complete lmerBayes block.
01|utilities-lmerBayes.r|llike.fixef.lmer|Strong|Not listed individually; inherits the explicit lmerBayes block recommendation.
01|utilities-lmerBayes.r|llike.model.lmer|Strong|Explicit recommendation to deprecate the complete lmerBayes block.
01|utilities-lmerBayes.r|residual.llike.lmerBayes|Strong|Explicit recommendation to deprecate the complete lmerBayes block.
01|utilities-lmerBayes.r|badSD|Strong|Explicit recommendation to deprecate the complete lmerBayes block.
01|utilities-lmerBayes.r|arrangeParam.llike.2D|Strong|Explicit recommendation to deprecate the complete lmerBayes block.
01|utilities-lmerBayes.r|arrangeParam.Gibbs.2D|Strong|Explicit recommendation to deprecate the complete lmerBayes block.
01|utilities-lmerBayes.r|saveParamFile|Strong|Explicit recommendation to deprecate the complete lmerBayes block.
01|utilities-lmerBayes.r|restoreParamFile|Strong|Explicit recommendation to deprecate the complete lmerBayes block.
01|utilities-lmerBayes.r|resetParam|Strong|Explicit recommendation to deprecate the complete lmerBayes block.
01|utilities-lmerBayes.r|summaryMCMC|Strong|Explicit recommendation to deprecate the complete lmerBayes block.
01|utilities-lmerBayes.r|recalculate.lmerBayesllike|Strong|Explicit recommendation to deprecate the complete lmerBayes block.
01|utilities-lmerBayes.r|covTocorr|Strong|Explicit recommendation to deprecate the complete lmerBayes block.
01|utilities-modelBayes.r|modelBayes|Strong|Explicit recommendation to deprecate the complete modelBayes block.
01|utilities-modelBayes.r|residual.llike.modelBayes|Strong|Explicit recommendation to deprecate the complete modelBayes block.
01|utilities-modelBayes.r|summaryModelMCMC|Strong|Explicit recommendation to deprecate the complete modelBayes block.
01|utilities-statistics.r|regression.Bayes|After checks|The old review suspected other functions might depend on it.
01|utilities-statistics.r|Gibbs.regslope|After checks|The old review suspected other functions might depend on it.
01|utilities-statistics.r|Gibbs.regsigma|After checks|The old review suspected other functions might depend on it.
01|utilities-statistics.r|Gibbs.normalmean|After checks|The old review suspected other functions might depend on it.
01|utilities-statistics.r|Gibbs.normalvar|After checks|The old review suspected other functions might depend on it.
01|utilities-statistics.r|model.xy|After checks|The old review suspected other functions might depend on it.
01|utilities-statistics.r|arrangeParam.llike|After checks|Considered overly specific; check other uses before removal.
01|utilities-statistics.r|arrangeParam.Gibbs|After checks|Considered overly specific; check other uses before removal.
01|utilities-statistics.r|llike.GaussModel|After checks|The old review suspected other functions might depend on it.
01|utilities-statistics.r|llike.GaussModelSD|After checks|The old review suspected other functions might depend on it.
01|utilities-statistics.r|BadParam|After checks|Considered overly specific; check other uses before removal.
01|utilities-statistics.r|graph.modeldiag|After checks|The old review suspected other functions might depend on it.
01|utilities-statistics.r|metrop1step|After checks|The old review suspected other functions might depend on it.
01|utilities-statistics.r|metrop1step.discrete|After checks|The old review suspected other functions might depend on it.
01|utilities-statistics.r|testmcmcfunc|After checks|The old review suspected other functions might depend on it.
05|demogchange-abund.fit.CTFS.r|model.littleR.Gibbs|Strong|Paper-specific model; explicitly recommended for deprecation.
05|demogchange-abund.fit.CTFS.r|full.abundmodel.llike|Strong|Paper-specific model; explicitly recommended for deprecation.
05|demogchange-abund.fit.CTFS.r|prob.N1|Strong|Paper-specific model; explicitly recommended for deprecation.
05|demogchange-abund.fit.CTFS.r|spmean.mort.abundGibbs|Strong|Paper-specific model; explicitly recommended for deprecation.
05|demogchange-abund.fit.CTFS.r|hyper.abundGibbs|Strong|Paper-specific model; explicitly recommended for deprecation.
05|demogchange-abund.fit.CTFS.r|hyper.mortGibbs|Strong|Paper-specific model; explicitly recommended for deprecation.
05|demogchange-abund.fit.CTFS.r|bad.asympower.param|Strong|Paper-specific model; explicitly recommended for deprecation.
05|demogchange-abund.fit.CTFS.r|bad.asymexp.param|Strong|Paper-specific model; explicitly recommended for deprecation.
05|demogchange-abund.fit.CTFS.r|fitSeveralAbundModel|Strong|Paper-specific model; explicitly recommended for deprecation.
05|demogchange-abund.fit.CTFS.r|graph.abundmodel|Strong|Paper-specific model; explicitly recommended for deprecation.
05|demogchange-abund.fit.CTFS.r|find.xaxis.hist|Weak|Low-priority exploratory graphic.
05|demogchange-abund.fit.CTFS.r|abundmodel.fit|Strong|Paper-specific model; explicitly recommended for deprecation.
05|demogchange-demogChange.r|lmerMortLinear|Weak|Low-priority model function; no firm deprecation decision recorded.
05|demogchange-demogChange.r|lmerMortFixedTime|Weak|Low-priority model function; no firm deprecation decision recorded.
05|growth-growthfit.bin.r|extract.growthdata|Weak|The old recommendation was to try to deprecate the growth-bin block.
05|growth-growthfit.bin.r|run.growthbin.manyspp|Weak|The old recommendation was to try to deprecate the growth-bin block.
05|growth-growthfit.bin.r|run.growthfit.bin|Weak|The old recommendation was to try to deprecate the growth-bin block.
05|growth-growthfit.bin.r|growth.flexbin|Weak|The old recommendation was to try to deprecate the growth-bin block.
05|growth-growthfit.bin.r|llike.linearbin.optim|Weak|The old recommendation was to try to deprecate the growth-bin block.
05|growth-growthfit.bin.r|defineBinBreaks|Weak|The old recommendation was to try to deprecate the growth-bin block.
05|growth-growthfit.bin.r|defineSDpar|Weak|The old recommendation was to try to deprecate the growth-bin block.
05|growth-growthfit.bin.r|enoughSamplePerBin|Weak|The old recommendation was to try to deprecate the growth-bin block.
05|growth-growthfit.bin.r|wideEnoughBins|Weak|The old recommendation was to try to deprecate the growth-bin block.
05|growth-growthfit.bin.r|bad.binparam|Weak|The old recommendation was to try to deprecate the growth-bin block.
05|growth-growthfit.bin.r|bad.binsdpar|Weak|The old recommendation was to try to deprecate the growth-bin block.
05|growth-growthfit.bin.r|calculateBinModel.BIC|Weak|The old recommendation was to try to deprecate the growth-bin block.
05|growth-growthfit.bin.r|calculateBinModel.DIC|Weak|Not listed individually; inherits the growth-bin block recommendation.
05|growth-growthfit.bin.r|calculateBinModel.AIC|Weak|The old recommendation was to try to deprecate the growth-bin block.
05|growth-growthfit.bin.r|calculateBinModel.bestpred|Weak|The old recommendation was to try to deprecate the growth-bin block.
05|growth-growthfit.bin.r|assembleBinOutput|Weak|Try to deprecate; reconsider binning separately in a more general context.
05|growth-growthfit.graph.r|graph.growthmodel.spp|Weak|Low-priority graphic associated with the growth-bin models.
05|growth-growthfit.graph.r|graph.growthmodel|Weak|Low-priority graphic associated with the growth-bin models.
05|growth-growthfit.graph.r|overlay.growthbinmodel|Weak|Low-priority graphic associated with the growth-bin models.
05|growth-growthfit.graph.r|compare.growthbinmodel|Weak|Low-priority graphic associated with the growth-bin models.
05|growth-growthfit.graph.r|graph.outliers.spp|Weak|Low-priority graphic associated with the growth-bin models.
05|growth-growthfit.graph.r|graph.outliers|Weak|Low-priority graphic associated with the growth-bin models.
05|growth-growthfit.graph.r|binGraphSampleSpecies|Weak|Low-priority graphic associated with the growth-bin models.
05|growth-growthfit.graph.r|binGraphManySpecies.Panel|Weak|Low-priority graphic associated with the growth-bin models.
05|growth-growthfit.graph.r|binGraphManySpecies|Weak|Low-priority graphic associated with the growth-bin models.",
  sep = "|",
  quote = "",
  colClasses = "character",
  check.names = FALSE
)

candidate_functions <- candidate_table$Function
candidate_table_display <- candidate_table
names(candidate_table_display)[c(1, 4, 5)] <- c(
  "Mod.", "Old rating", "Old evaluation summary"
)

candidate_table_latex <- knitr::kable(
  candidate_table_display,
  format = "latex",
  booktabs = TRUE,
  longtable = TRUE,
  row.names = FALSE
)

candidate_table_latex <- sub(
  "\\begin{longtable}{lllll}",
  paste0(
    "\\begin{longtable}{@{}",
    "L{0.9cm}L{3.2cm}L{5cm}L{2.4cm}L{8.8cm}",
    "@{}}"
  ),
  candidate_table_latex,
  fixed = TRUE
)

cat(
  "\\begin{landscape}\n",
  "\\small\n",
  "\\setlength{\\tabcolsep}{4pt}\n",
  candidate_table_latex,
  "\n\\end{landscape}\n",
  sep = ""
)
```

Some formal tests indicate that these functions, if conserved, would require maintenance and debugging. In other words: conservation will come with overhead costs.

-   The [CTFS development notes](https://github.com/forestgeo/ctfs/blob/cc32c7c2a3959ef60b13144f956e99d3a0515322/NEWS.md) record that `modelBayes()` still entered debugging and failed after an attempted maintenance fix.
-   `growth.flexbin()`, `run.growthfit.bin()`, and `run.growthbin.manyspp()` failed in the growth tutorial. Their attempted calls remain commented out in [test-growth_vs_dbh.R](https://github.com/forestgeo/ctfs/blob/cc32c7c2a3959ef60b13144f956e99d3a0515322/tests/testthat/test-growth_vs_dbh.R).
-   The `lmerBayes()` regression test was disabled because its required data were no longer available. Even when enabled, it checked only the names of the returned components, not the fitted result; see [test-replace_subset.R](https://github.com/forestgeo/ctfs/blob/cc32c7c2a3959ef60b13144f956e99d3a0515322/tests/testthat/test-replace_subset.R).
-   The development notes report that one test of `model.littleR.Gibbs()` and `fitSeveralAbundModel()` took approximately 20 minutes. No active automated test for that model remains in the CTFS test suite.

# Due diligence for dependents and dependencies

This due diligence has two main goals:

1.  identify any function outside the candidate set that would break if a candidate were removed; and
2.  identify non-candidate functions that may no longer be necessary if all candidates are removed.

This is a static analysis of calls among top-level fgeo2 functions. It does not detect calls constructed with mechanisms such as `get()`, `do.call()`, or `match.fun()`, or uses from tutorials, tests, and external user code. Any apparent absence of dependents therefore remains a prompt for final source and test checks, not automatic authorization to remove code.

## Pre-work snapshot

The dependency structure below uses every R script in the `R/new` directory of all seven fgeo2 modules. The exact pre-work snapshot is commit [`040b73e`](https://github.com/gabrielareto/fgeo2/commit/040b73e0cd9a0d55d64c1639b22159d5c6cbd20e). Git reads the files directly from that commit, so later changes in the working tree are not mixed into the evaluation.

```{r}

path <- normalizePath(system2("git", c("rev-parse", "--show-toplevel"),
                              stdout = TRUE)[1])

library(igraph, quietly = TRUE)
source(file.path(path, "evaluation/0-code-to-source.r"))

pre_work_commit <- "040b73e0cd9a0d55d64c1639b22159d5c6cbd20e"

modules <- c(
  "01_utilities", "02_biomass", "03_spatial",
  "04_abundance_diversity", "05_demography_growth",
  "06_topography_habitat", "07_mapping_plotting"
)

all_files <- system2(
  "git",
  c("-C", shQuote(path), "ls-tree", "-r", "--name-only", pre_work_commit),
  stdout = TRUE
)

module_pattern <- paste(modules, collapse = "|")
scripts <- all_files[
  grepl(paste0("^(", module_pattern, ")/R/new/.*\\.[Rr]$"), all_files)
]

code_env <- new.env(parent = .GlobalEnv)

for(file in scripts)
{
  code <- system2(
    "git",
    c("-C", shQuote(path), "show", paste0(pre_work_commit, ":", file)),
    stdout = TRUE
  )

  eval(parse(text = paste(code, collapse = "\n")), envir = code_env)
}

objects <- mget(
  ls(code_env, all.names = TRUE),
  envir = code_env,
  inherits = FALSE
)

all_functions <- names(objects)[vapply(objects, is.function, logical(1))]

edges <- get_all_dependencies(envir = code_env)
edges <- edges[, c("from_function", "to_function"), drop = FALSE]
colnames(edges) <- c("from", "to")

candidates <- intersect(candidate_functions, all_functions)
missing_candidates <- setdiff(candidate_functions, all_functions)

if(length(missing_candidates))
  stop("Candidate functions not found: ", paste(missing_candidates, collapse = ", "))

```

## Dependents that could break

For each candidate, the following code follows dependency arrows through all levels and retains only dependents outside the candidate set. These are the functions that would require attention before removal.

```{r}
#| results: asis

dependent_rows <- list()

for(candidate in candidates)
{
  direct_dependents <- unique(edges$to[edges$from == candidate])
  direct_dependents <- setdiff(direct_dependents, candidates)

  all_dependents <- downstream_dependents(candidate, edges)
  all_dependents <- setdiff(all_dependents, candidates)

  if(length(all_dependents) == 0)
    next

  dependent_rows[[length(dependent_rows) + 1L]] <- data.frame(
    Candidate = candidate,
    `Direct dependents outside candidates` = paste(direct_dependents, collapse = ", "),
    `Dependents outside candidates at any level` = paste(all_dependents, collapse = ", "),
    check.names = FALSE
  )
}

if(length(dependent_rows) == 0)
{
  cat("**Result:** No function outside the candidate set depends on a candidate in the pre-work snapshot.\n")
} else {
  dependent_table <- do.call(rbind, dependent_rows)
  dependent_table_latex <- knitr::kable(
    dependent_table,
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE,
    row.names = FALSE
  )

  dependent_table_latex <- sub(
    "\\begin{longtable}{lll}",
    "\\begin{longtable}{@{}L{4cm}L{9cm}L{9cm}@{}}",
    dependent_table_latex,
    fixed = TRUE
  )

  cat(
    "\\begin{landscape}\n",
    "\\footnotesize\n",
    "\\setlength{\\tabcolsep}{4pt}\n",
    dependent_table_latex,
    "\n\\end{landscape}\n",
    sep = ""
  )
}

```

## Dependencies that may become unnecessary

This check follows dependencies upstream through all levels. A non-candidate dependency is marked as potentially unnecessary when it belongs to the candidate dependency chain and has no downstream use outside that potential removal set.

```{r}
#| results: asis

candidate_dependencies <- character()

for(candidate in candidates)
  candidate_dependencies <- c(
    candidate_dependencies,
    upstream_dependencies(candidate, edges)
  )

candidate_dependencies <- unique(candidate_dependencies)
candidate_dependencies <- setdiff(candidate_dependencies, candidates)
potential_removal_set <- c(candidates, candidate_dependencies)

dependency_rows <- list()

for(dependency in candidate_dependencies)
{
  all_users <- downstream_dependents(dependency, edges)
  candidate_users <- intersect(all_users, candidates)
  other_users <- setdiff(all_users, potential_removal_set)

  conclusion <- "Still required elsewhere"
  if(length(other_users) == 0)
    conclusion <- "Potentially unnecessary"

  dependency_rows[[length(dependency_rows) + 1L]] <- data.frame(
    Dependency = dependency,
    `Candidate functions downstream` = paste(candidate_users, collapse = ", "),
    `Other downstream functions` = if(length(other_users)) paste(other_users, collapse = ", ") else "None detected",
    Conclusion = conclusion,
    check.names = FALSE
  )
}

if(length(dependency_rows) == 0)
{
  cat("**Result:** No non-candidate fgeo2 dependencies were detected.\n")
} else {
  dependency_table <- do.call(rbind, dependency_rows)
  dependency_table <- dependency_table[
    order(
      dependency_table$Conclusion != "Potentially unnecessary",
      dependency_table$Dependency
    ),
  ]

  dependency_table_latex <- knitr::kable(
    dependency_table,
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE,
    row.names = FALSE
  )

  dependency_table_latex <- sub(
    "\\begin{longtable}{llll}",
    "\\begin{longtable}{@{}L{3.4cm}L{8.1cm}L{8.1cm}L{2.5cm}@{}}",
    dependency_table_latex,
    fixed = TRUE
  )

  cat(
    "\\begin{landscape}\n",
    "\\footnotesize\n",
    "\\setlength{\\tabcolsep}{4pt}\n",
    dependency_table_latex,
    "\n\\end{landscape}\n",
    sep = ""
  )
}

```

# Current evaluation and decisions

I have reviewed all functions one by one, and in the context of their corresponding working block.

As far as I understand, all this functionality is better covered by standard applications of WinBugs, STAN, `brms` package in R, and other standard tools for Bayesian / hierarchical modeling.

The modeling task itself requires expertise. Experts do know these tools. Overall, it seems very unlikely that any member of the ForestGEO / ATFS communities use (or should use) the code developed by Condit and stored in CTFS.

The recent ForestGEO workshop in Krabi (Thailand) suggests that these tools are becoming more and more accessible and easy to use, even for the average user. The AI revolution will probably make Bayesian statistics even easier to implement.

Some of Condit's code seems legacy code with the partial goal of reproducibility of results, particularly the code in the "abund.fit" block. Although reproducibility is a valuable goal, there are other ways of ensuring that. It does not seem useful that fgeo2 functions as a repository for legacy code.

All points in the same direction: a massive, direct removal of all the functions mentioned above is justified, and it would not have downstream consequences. All code remains under "old" folders as part of the fgeo2 distribution, though.

# Package structure after removals

The graph is intentionally deferred until after the removals. Once the work is complete, replace the post-work commit and date below and enable the chunk. The graph will use every top-level function in the seven `R/new` directories, including functions with no detected internal dependency edges.

```{r}
#| eval: false
#| out-width: "100%"

post_work_commit <- "REPLACE_WITH_POST_WORK_COMMIT"
post_work_date <- as.Date("YYYY-MM-DD")

post_all_files <- system2(
  "git",
  c("-C", shQuote(path), "ls-tree", "-r", "--name-only", post_work_commit),
  stdout = TRUE
)

post_scripts <- post_all_files[
  grepl(paste0("^(", module_pattern, ")/R/new/.*\\.[Rr]$"), post_all_files)
]

post_code_env <- new.env(parent = .GlobalEnv)

for(file in post_scripts)
{
  code <- system2(
    "git",
    c("-C", shQuote(path), "show", paste0(post_work_commit, ":", file)),
    stdout = TRUE
  )

  eval(parse(text = paste(code, collapse = "\n")), envir = post_code_env)
}

post_objects <- mget(
  ls(post_code_env, all.names = TRUE),
  envir = post_code_env,
  inherits = FALSE
)

post_functions <- names(post_objects)[vapply(post_objects, is.function, logical(1))]

post_edges <- get_all_dependencies(envir = post_code_env)
post_edges <- post_edges[, c("from_function", "to_function"), drop = FALSE]
colnames(post_edges) <- c("from", "to")

post_vertices <- data.frame(name = post_functions)
g.after <- graph_from_data_frame(
  post_edges,
  directed = TRUE,
  vertices = post_vertices
)

remaining_candidates <- intersect(candidate_functions, post_functions)
post_downstream <- character()

for(candidate in remaining_candidates)
  post_downstream <- c(post_downstream, downstream_dependents(candidate, post_edges))

post_downstream <- unique(post_downstream)
post_downstream <- setdiff(post_downstream, remaining_candidates)

V(g.after)$color <- "gray80"
V(g.after)$color[V(g.after)$name %in% post_downstream] <- "yellow"
V(g.after)$color[V(g.after)$name %in% remaining_candidates] <- "orange"

post_work_pdf <- file.path(
  path,
  "evaluation/results",
  paste0(
    "fgeo2-dependencies-", substr(post_work_commit, 1, 7), "-",
    format(post_work_date, "%Y-%m-%d"), ".pdf"
  )
)

dir.create(dirname(post_work_pdf), recursive = TRUE, showWarnings = FALSE)

set.seed(1)

pdf(post_work_pdf, height = 33, width = 33)
par(mar = c(0, 0, 2, 0))

plot(
  g.after,
  layout = layout.fruchterman.reingold,
  vertex.label = V(g.after)$name,
  vertex.size = 6,
  vertex.color = V(g.after)$color,
  vertex.frame.color = "white",
  vertex.label.color = "black",
  vertex.label.cex = 2,
  edge.color = "gray45",
  edge.width = 0.7,
  edge.arrow.size = 0.3
)

legend(
  "topleft",
  legend = c(
    "Candidate remaining",
    "Downstream dependent",
    "Other function"
  ),
  pt.bg = c("orange", "yellow", "gray80"),
  pch = 21,
  pt.cex = 2,
  bty = "n"
)

dev.off()

post_work_pdf_relative <- file.path("results", basename(post_work_pdf))
knitr::include_graphics(post_work_pdf_relative)

```
