---
title: "Recovering coordinates from paper maps"
author: "Gabriel Arellano"
date: today
format:
  pdf:
    code-overflow: wrap
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
editor: visual
---

# Purpose

Points clicked on a scanned or photographed map are recorded in image coordinates. Control points connect that image to known plot coordinates. The recommended transformation is a **homography**, using the four corners of the mapped quadrat. It corrects photographic perspective and performed almost as well as the affine transformation on flatbed scans in the [coordinate-recovery evaluation](../../evaluation/5-imagej-and-geometry.qmd). Use an **affine transformation** as a fallback for a flat or nearly flat image when one corner is missing or a corner click is unusually unreliable.

These transformations correct one whole planar image. They do not repair folds, tears, uneven stretching or errors in the original field map.

# Setup

Run this tutorial inside a clone of the fgeo2 repository. The code uses only base R and the example data stored in the repository.

```{r}
#| label: setup

p <- system2("git", c("rev-parse", "--show-toplevel"), stdout = TRUE)
if(!is.null(attr(p, "status")) || length(p) != 1L)
	stop("Run this tutorial inside the fgeo2 Git repository.")
p <- normalizePath(p)

source(file.path(
	p, "06_topography_habitat", "R", "new", "georeference.R"
))

example.path <- file.path(
	p, "data_examples", "coordinate_recovery_test"
)
```

# Recommended method: homography

Case A is an untreated paper map digitized with a flatbed scanner. The file contains four clicked corners followed by clicked test points. The four known corners describe a 10 × 10 m quadrat in global plot coordinates.

```{r}
#| label: prepare-example

clicked <- read.csv(file.path(example.path, "points test A.csv"))

known.controls <- data.frame(
	id = paste0("corner_", LETTERS[1:4]),
	x = c(50, 50, 60, 60),
	y = c(50, 60, 60, 50)
)

# Match controls by identifier; do not rely on file row order.
control.order <- match(known.controls$id, clicked$id)
stopifnot(!anyNA(control.order))

known.xy.of.controls <- as.matrix(known.controls[, c("x", "y")])
clicked.xy.of.controls <- as.matrix(
	clicked[control.order, c("x", "y")]
)
rownames(known.xy.of.controls) <- known.controls$id
rownames(clicked.xy.of.controls) <- clicked$id[control.order]
stopifnot(identical(
	rownames(known.xy.of.controls),
	rownames(clicked.xy.of.controls)
))

is.control <- clicked$id %in% known.controls$id
clicked.xy.of.points <- as.matrix(clicked[!is.control, c("x", "y")])
rownames(clicked.xy.of.points) <- clicked$id[!is.control]
```

The result uses the units and coordinate system of `known.xy.of.controls`. Here it is returned directly in global plot coordinates and metres.

```{r}
#| label: apply-homography

homography.xy <- apply_homography_transformation(
	known.xy.of.controls = known.xy.of.controls,
	clicked.xy.of.controls = clicked.xy.of.controls,
	clicked.xy.of.points = clicked.xy.of.points
)

recovered <- data.frame(
	id = rownames(homography.xy),
	global.x = homography.xy[, "x"],
	global.y = homography.xy[, "y"],
	row.names = NULL
)

head(recovered)
```

The example was constructed from known points, so it also provides a simple check of the result.

```{r}
#| label: check-example

known.points <- expand.grid(
	x = seq(48, 62, by = 2),
	y = seq(48, 62, by = 2)
)
known.points$id <- paste0("p", seq_len(nrow(known.points)))
known.order <- match(recovered$id, known.points$id)
stopifnot(!anyNA(known.order))

error.cm <- 100 * sqrt(
	(recovered$global.x - known.points$x[known.order])^2 +
	(recovered$global.y - known.points$y[known.order])^2
)

round(c(
	mean.error.cm = mean(error.cm),
	median.error.cm = median(error.cm),
	maximum.error.cm = max(error.cm)
), 1)
```

```{r}
#| label: fig-recovered-points
#| fig-cap: "Coordinates recovered from test case A. The square marks the four control locations."
#| fig-width: 5
#| fig-height: 5

plot(
	recovered$global.x,
	recovered$global.y,
	asp = 1,
	xlab = "Global plot x (m)",
	ylab = "Global plot y (m)",
	pch = 20
)
polygon(known.controls$x, known.controls$y, border = "red", lwd = 2)
```

# Fallback: affine transformation

An affine transformation needs at least three non-collinear controls. It is a practical fallback for a flatbed scan when one corner is unavailable. Do not use three controls to correct appreciable photographic perspective. If four or more reliable controls are available, retain them: the affine fit uses all of them by least squares.

This example deliberately omits corner D from the same scan.

```{r}
#| label: apply-affine

available.controls <- paste0("corner_", LETTERS[1:3])
available <- match(available.controls, known.controls$id)

affine.xy <- apply_affine_transformation(
	known.xy.of.controls = known.xy.of.controls[available, ],
	clicked.xy.of.controls = clicked.xy.of.controls[available, ],
	clicked.xy.of.points = clicked.xy.of.points
)

affine.recovered <- data.frame(
	id = rownames(affine.xy),
	global.x = affine.xy[, "x"],
	global.y = affine.xy[, "y"],
	row.names = NULL
)

head(affine.recovered)
```

# Obtaining clicked coordinates

The transformations do not depend on ImageJ. Any tool that exports matching point identifiers and image x-y coordinates can be used. In R, the minimal interactive approach is `locator()`; see the complete [example script](../../evaluation/coordinate-locator.R). Click the four corners in a stated order, then the mapped points, and press Esc or **Finish** when done. The direction of the image y-axis does not matter if every corner is identified correctly.

```{r}
#| label: click-points
#| eval: false
#| purl: false

pdf.path <- file.path(example.path, "test A.pdf")
map.image <- magick::image_read(pdf.path, density = "300x300")[1]
map.raster <- as.raster(map.image)
plot(map.raster, interpolate = FALSE)

clicked <- locator(n = 1000, type = "p", pch = 19, col = "red")
clicked <- data.frame(x = clicked$x, y = clicked$y)
clicked$id <- c(
	paste0("corner_", LETTERS[1:4]),
	paste0("point_", seq_len(nrow(clicked) - 4L))
)
clicked <- clicked[, c("id", "x", "y")]
write.csv(clicked, "clicked-points.csv", row.names = FALSE)
```

ImageJ or [Fiji](https://imagej.net/software/fiji/downloads) can also export the points. The historical workflow used a modified [PointPicker plugin](https://ctfs.si.edu/Public/DatabaseSetup/Resources/imageJ/pointpicker-modified3.zip). Point collection and coordinate recovery are separate steps; ImageJ is not required for either transformation above.

# Local and global plot coordinates

The known controls define the output coordinates. Corners such as `(50, 50)`, `(50, 60)`, `(60, 60)` and `(60, 50)` return global plot coordinates directly. Corners entered as `(0, 0)`, `(0, 10)`, `(10, 10)` and `(10, 0)` return coordinates local to that quadrat.

If a process must use local controls, convert its results afterwards with `compile_points_from_quadrats()`. The function and a fuller explanation are in the [quadrat source](../../03_spatial/R/new/quadrats.R) and [quadrat tutorial](../../03_spatial/tutorials/quadrats.qmd).

```{r}
#| label: local-to-global

source(file.path(p, "03_spatial", "R", "new", "quadrats.R"))

local.points <- data.frame(
	id = c("tree_1", "tree_2"),
	x = c(2.1, 8.4),
	y = c(3.7, 6.2)
)

quadrat.locations <- data.frame(
	quadrat_id = "q_50_50",
	xmin = 50,
	ymin = 50,
	xmax = 60,
	ymax = 60
)

global.points <- compile_points_from_quadrats(
	inner.x = local.points$x,
	inner.y = local.points$y,
	point.id = local.points$id,
	quadrat.id = rep("q_50_50", nrow(local.points)),
	quadrat.locations = quadrat.locations
)

global.points
```

These are plot-coordinate transformations, not conversions between geographic coordinate reference systems.

# Historical CTFS compatibility only

The CTFS/ImageJ conversion is not the recommended pipeline and will not be retained in fgeo2. The archived [`imageJ.r` implementation](../../evaluation/Condit/topography/imageJ.r) remains available with its [geometry dependencies](../../evaluation/Condit/utilities/geometry.r). The original workflow is documented in the 2011 CTFS [PDF tutorial](https://ctfs.si.edu/ctfsdev/CTFSRPackageNew/files/tutorials/imageJTutorial/imageJ.pdf) and [HTML tutorial](https://ctfs.si.edu/ctfsdev/CTFSRPackageNew/index.php/web/tutorials/imageJTutorial/index.html).

The archived `distance.to.side()` passes one-column data frames instead of numeric intersection coordinates. If the old workflow is still required, run the following session-only replacement after loading the archived CTFS scripts. It fixes that one bug only. It does not fix inconsistent corner defaults, exterior points folded into a quadrat, or other limitations of the legacy workflow. Supply all four corner labels explicitly when calling the old functions.

```{r}
#| label: legacy-ctfs-fix
#| eval: false
#| purl: false

condit.path <- file.path(p, "evaluation", "Condit")
source(file.path(condit.path, "startCTFS", "startCTFS.r"))
startCTFS(condit.path)

distance.to.side.original <- distance.to.side

distance.to.side <- function(pt, target, side1, side2)
{
	perpendicular <- perpendicular.distance(
		b = target[1], m = target[2], x = pt[1], y = pt[2]
	)

	parallel <- drop(as.matrix(parallel.line(
		b = side1[1], m = side1[2], x = pt[1], y = pt[2]
	)))
	intersection <- intersection.of.lines(
		b1 = parallel[1], m1 = parallel[2],
		b2 = target[1], m2 = target[2]
	)
	distance1 <- xydist(pt[1], pt[2], intersection$x, intersection$y)

	parallel <- drop(as.matrix(parallel.line(
		b = side2[1], m = side2[2], x = pt[1], y = pt[2]
	)))
	intersection <- intersection.of.lines(
		b1 = parallel[1], m1 = parallel[2],
		b2 = target[1], m2 = target[2]
	)
	distance2 <- xydist(pt[1], pt[2], intersection$x, intersection$y)

	distances <- c(perpendicular, distance1, distance2)
	c(mean(distances), sd(distances) / mean(distances))
}

stopifnot(isTRUE(all.equal(
	unname(distance.to.side(
		pt = c(5, 5),
		target = c(0, Inf),
		side1 = c(10, 0),
		side2 = c(0, 0)
	)),
	c(5, 0)
)))
```
